---
title: "More on Survival"
author: "Erin Feichtinger"
date: "Monday, February 29, 2016"
output: html_document
---

###Review on FSJ Survival Analyses so far
Alright so I figured out how the data need to be set up for the model to work correctly. I have fit both Cox proportional hazard and accelerated life models to 2 subsets of the scrub-jay data.  
###Subset 1: All known age birds from 1981 to 2015
The first subset of data consists of all known age birds who fledged between 1981 and 2015. The data frame includes jay ID (both band number and color), date fledged, date last seen, sex and a few other fields. There are 1076 females and 1013 males for a total sample size of 2089. 
```{r, echo=FALSE}
setwd("C:/Users/Erin/LittleBlueDinos")
library(survival)
library(car)
library(kinship2)
library(SurvRegCensCov)
#read in CSV of all known-age birds 
birds <- read.csv("Erin_Surv_All.csv")
#str(birds)

#remove duplicates 
birds2 <- birds[!duplicated(birds),]
#str(birds2)

colnames(birds2)[7] <- "LastObsDate"

#convert dates to date format
birds2$FldgDate <- as.Date(birds2$FldgDate, format = "%m/%d/%Y")
birds2$LastObsDate <- as.Date(birds2$LastObsDate, format = "%m/%d/%Y")

#subtract dates to get number of days
date.diff<- birds2$LastObsDate-birds2$FldgDate

#add to data frame - survival period in days
birds2["days"] <- date.diff
birds2$days <-as.numeric(birds2$days)
#and survival period in years 
birds2["yrs"] <- birds2$days/365.25

#very important piece of code for the model to work properly
#remove any zero or negative values in days and years 
birds2 <- subset(birds2, birds2$days > 0 & birds2$yrs > 0)

#add column for censorship status 
birds2["censorship"] <- 1
#If last observed date = 10/14/2015, 0 for still alive today
birds2$censorship[which(birds2$LastObsDate=="2015-10-14")]<-0


year <- as.POSIXlt(birds2$FldgDate)$year+1900
birds2["FYear"] <- year

str(birds2)
#change back to numeric for survival object 
birds2$FldgDate <- as.numeric(birds2$FldgDate)
birds2$LastObsDate <- as.numeric(birds2$LastObsDate)
birds2$days <- as.numeric(birds2$days)
birds2$yrs <- as.numeric(birds2$yrs)
birds2$FYear <- as.factor(birds2$FYear)

#Create survival object based off Gordon's Cactus Finch example
survobj <- Surv(birds2$yrs, birds2$censorship, type =c('right'))

all.lifetab <- survfit(survobj~1)
all.log <- plot(all.lifetab, log = "y", ylim=c(0.001,2),
     xlab =  "Time (years)", ylab = "Cumulative Survival", 
      main = "Survival of Known Age Birds - Log Scale")

all.sex <- survfit(survobj~birds2$Sex, conf.type = "log-log")
all.logsex <- plot(all.sex, col = c("blue", "red"), log = "y", 
            ylim = c(0.001,2), xlab = "Time (years)",
    ylab = "Cumulative Survival", main = "Survival by Sex - Log Scale")
legend("topright", c("Females","Males"), col = c("blue","red"),
       lty = c(1,2), lwd = 2)

```

Both figures represent Kaplan-Meier estimates of survival over time. Here, survival is on a natural log scale. The first survivorship curve with both sexes combined has condfidence intervals displayed. The survivorship curves for each sex do not. We can see that the lines for males and females are indistinguishable at first then just before year 2 the survival estimates for females is lower. However, we need to fit some models to know if male and female survival are truly different. First, I started with Cox Proportional Hazard models. These models focus on the hazard function, or the instantaneous risk of mortality. The sign of the estimated coefficients in the model tell us whether a covariate increases or decreases hazard.   

```{r}
#Cox PH Model with sex as predictor 
cox <- coxph(survobj ~ birds2$Sex, data = birds2)
summary(cox)
```
We can see from the output that sex is a significant predictor, however, the R^2=0.008. The sign for the coefficient is negative meaning that being male reduces your hazard or risk of dying at any given time. But again, sex does not explain much of the variation in hazard over time. We can add some predictors including year of fledging and group size.  

Another type of model for time to event or survival data are accelerated life models or accelerated failure time models. Let's look at the fit of the AFT model with sex as a predictor. 
```{r}
AFT.sex <- survreg(survobj ~ Sex, data = birds2, dist = "weibull")
summary(AFT.sex)
```
The output here shows that sex is significant but I'm not sure it means anything. Notice that the sign of the estimated coefficient for SexM is positive. This is the opposite of the Cox model and it should be! AFT models focus on survival, or time to event, rather than focus directly on the hazard function. For some individuals, death will come sooner (shorter failure time) than others. A positive coefficient means that the covariate increases failure time, or increaes survival time. The effect of sex on survival is consistent between the two types of models. 

###Subset 2: Breeders from 1981 to 2015
The second subset of data are birds of known and unknown age who bred in the population. The first date of breeding, or start of the interval, is the minimum date a bird was known to lay or sire an egg. The end of the interval is date last seen or October 14, 2015. 
```{r, echo=FALSE}
bird.df <- read.csv("Breeders_Fall.csv")
str(bird.df)

#remove duplicates  - for years where there was more than one nest in a year
jay.df<- bird.df[!duplicated(bird.df),]
str(jay.df)

#colnames(jay.df)[1] <- "ID"
#colnames(jay.df)[2] <- "Band"
colnames(jay.df)[5] <- "MinDate"

#convert dates to date format
jay.df$MinDate <- as.Date(jay.df$MinDate, format = "%m/%d/%Y")
jay.df$LastObsDate <- as.Date(jay.df$LastObsDate, format = "%m/%d/%Y")

#subtract dates to get number of days
date.diff<- jay.df$LastObsDate-jay.df$MinDate

#and survival period in years, account for leap year 
jay.df["Yrs"] <- as.numeric(date.diff/365.25)

jay.df$FirstYr <- as.factor(jay.df$FirstYr)

#very important piece of code for the model to work properly, remove any 
#weird entries like birds that have negative years of experience or a negative
#survival interval 
jay.df <- subset(jay.df, jay.df$Days > 0)


#add column for censorship status, in survival package - 0=alive, 1=dead
jay.df["censorship"] <- 1
#If last observed date = 10/14/2015, 0 for still alive today
jay.df$censorship[which(jay.df$LastObsDate=="2015-10-14")]<-0

#Check for correct structure
str(jay.df)

#How many males and females?
sum(jay.df$Sex == "M")
sum(jay.df$Sex == "F")

#change back to numeric for survival object 
jay.df$MinDate <- as.numeric(jay.df$MinDate)
jay.df$LastObsDate <- as.numeric(jay.df$LastObsDate)

#Create survival object - IS THIS CORRECT?? 
jay.ob <- Surv(jay.df$Yrs, jay.df$censorship, type =c('right'))
jay.lifetab <- survfit(jay.ob~1, conf.type = "log-log")

jay.fitlog <- plot(jay.lifetab, xlab = "Time (years)",
                   log = "y", ylim = c(0.001,2), ylab = "Cumulative Survival", 
                   main = "FL Scrub Breeder survival Log Scale")

km.sex <- survfit(jay.ob ~ jay.df$Sex, conf.type = "log-log")
sex.log <- plot(km.sex, col = c("navy","red"), log = "y", ylim = c(0.001,2),
          lty  = c(1,2), xlab = "Time (years)",ylab = "Cumulative Survival", 
                main = "Survival by Sex Log Scale")
legend("topright", c("Females","Males"), col = c("navy","red"), 
       lty = c(1,2), lwd =1)

```
The survival curves are the Kaplan-Meier estimates of survival shown for both sexes combined and separately. The curves for male and female breeders overlap for about 8 years from the start. Note that time 0 is when an individual started breeding. The average age at first breeding for males and females is 2.5 years. 

Here are the curves again so we can compare the estimates between the subsets of data, all known age birds and breeders from 1981 to 2015. 
```{r, echo=FALSE}
all.log <- plot(all.lifetab, log = "y", ylim=c(0.001,2),
     xlab =  "Time (years)", ylab = "Cumulative Survival", 
      main = "Survival of Known Age Birds - Log Scale")

jay.fitlog <- plot(jay.lifetab, xlab = "Time (years)",
                   log = "y", ylim = c(0.001,2), ylab = "Cumulative Survival", 
                   main = "FL Scrub Breeder survival Log Scale")

all.logsex <- plot(all.sex, col = c("blue", "red"), log = "y", 
            ylim = c(0.001,2), xlab = "Time (years)",
    ylab = "All Birds Cumulative Survival", main = "Survival by Sex - Log Scale")
legend("topright", c("Females","Males"), col = c("blue","red"),
       lty = c(1,2), lwd = 2)

sex.log <- plot(km.sex, col = c("navy","red"), log = "y", ylim = c(0.001,2),lty  = c(1,2), xlab = "Time (years)",ylab = "Cumulative Survival", 
      main = "Breeder Survival by Sex Log Scale")
legend("topright", c("Females","Males"), col = c("navy","red"), 
       lty = c(1,2), lwd =1)

```
###Cox Proportional Hazard models

```{r}
#Age at first breeding 
cox.age <- coxph(jay.ob ~ AgeFirstBreed, data = jay.df)
summary(cox.age)
```

From the output we can see that the estimated coefficient for age at first time breeding has a significant z statistic. The sign is positive so increasing age at first breeding increases hazard. However, look at the R^2. It is 0.007, so that suggests not much is explained by age at first breeding. 

The next model has both sex and age at first breeding as covariates.

```{r}
cox.sage <- coxph(jay.ob ~ AgeFirstBreed + Sex, data = jay.df)
summary(cox.sage)
```

Age at first breeding has a significant z score but sex does not. R^2 = 0.008. 

One more to try is year of first breeding. 
```{r}
cox3 <- coxph(jay.ob ~ FirstYr, data = jay.df)

```

I'm not showing the output because it's long. I coded year as a factor so there are 34 levels, one for each year. The coefficients for most years are not significant. Is it correct to code the first year as a factor and to have as many levels as years and an estimate for each? 2011 - 2015 have significant z stats with a positive estimate for those coefficients, meaning an increased hazard. R^2 = 0.24, but I'm not sure that this is correct. 

```{r}
#All 3 covariates - age at first breeding, sex, and year of first breeding
#No interactions
cox4 <- coxph(jay.ob ~ AgeFirstBreed + Sex + FirstYr, data = jay.df)
summary(cox4)
```
Consistent with the models of each covariate alone, age is significant and so are some levels of year. R^2 = 0.23, which is about the same as the model with first year alone. 

Just to be thorough, let's look at age and first year as covariates.
```{r}
cox5 <- coxph(jay.ob ~ AgeFirstBreed + FirstYr, data = jay.df)
summary(cox5)
```
Same as before with regards to the coefficients and R^2 is very similar. 

```{r}
extractAIC(cox.age)
extractAIC(cox.sage)
extractAIC(cox3)
extractAIC(cox4)
extractAIC(cox5)
```
Models 4 and 5 have the lowest AIC scores of this group of models. The AIC scores are identical. Model 4 is age at first breeding, sex and year at first breeding. Model 5 is age and year. 

###Accelerated Life Models
```{r}
#AFT model with Weibull distribution and age at first breeding 
AFT.weibull <- survreg(jay.ob ~ AgeFirstBreed, data = jay.df, dist = "weibull")
summary(AFT.weibull)

#AFT model with Weibull distribution and sex
AFT.weibull2 <- survreg(jay.ob ~ Sex, data = jay.df, dist = "weibull")
summary(AFT.weibull2)

#AFT model with Weibull distribution and year of first breeding
AFT.weibull3 <- survreg(jay.ob ~ FirstYr, data = jay.df, dist = "weibull")
summary(AFT.weibull3)

#All 3 covariates 
AFT.weibull4 <- survreg(jay.ob ~ AgeFirstBreed + Sex + FirstYr, 
      data = jay.df, dist = "weibull")
summary(AFT.weibull4)
```
I don't know how to interpret the output for the AFT models and how to tell if they are "good" descriptions of the data. 

```{r}
extractAIC(AFT.weibull)
extractAIC(AFT.weibull2)
extractAIC(AFT.weibull3)
extractAIC(AFT.weibull4)
```

Model 4 is the most supported according to the AIC scores of the set of models compared. This model has age, year and sex as covariates. 

###The next steps
1. Get group size in April for all birds and breeders from the database. I found an existing query that gives the group size in April of each year by territory. I have to figure out how to connect the field of group size to each individual bird. 
2. Get subset of data for birds when they were heplers. 

