---
title: "FSJ Surv June Week 4"
author: "Erin Feichtinger"
date: "June 21, 2016"
output: html_document
---
```{r, echo = FALSE}

library(survival)
library(car)
library(coxme)
library(kinship2)
library(SurvRegCensCov)

## Read in April Census Dates
aprilD <- read.csv("April_Census_Dates.csv")
## Convert to date object
aprilD$CensusDate <- as.Date(aprilD$CensusDate, format = "%m/%d/%Y")
## First data table to deal with, known age birds to figure out who died in the first year
## Read in CSV file 
bird.df <- read.csv("Erin_June_FY.csv")
str(bird.df)
## 2371 individuals 
## Change column name to Days for days lived and add a year column
colnames(bird.df)[8] <- "Days"
bird.df["Yrs"] <- bird.df$Days/365.25
## convert dates to date format
bird.df$FldgDate <- as.Date(bird.df$FldgDate, format = "%m/%d/%Y")
bird.df$LastObsDate <- as.Date(bird.df$LastObsDate, format = "%m/%d/%Y")

bird.df["Censor"] <- 1
#Birds that are right censored (indicated by 0) are still alive after 1 year
#post fledgling 
bird.df$Censor[which(bird.df$Yrs > 1)]<-0
#subset to get rid of years less than 0
yrlg.df <- subset(bird.df, bird.df$Yrs > 0)

#change back to numeric for survival object 
yrlg.df$FldgDate <- as.numeric(yrlg.df$FldgDate)
yrlg.df$LastObsDate <- as.numeric(yrlg.df$LastObsDate)

#survival object 
yrlg.ob <- Surv(yrlg.df$Yrs, yrlg.df$Censor, type = c('right'))

my.fit <- survfit(yrlg.ob~1, conf.type = "log-log")

plot.fit <- plot(my.fit, xlab = "Time (years)",
      log = "y", ylim = c(0.4, 1),xlim=c(0,1), ylab = "Cumulative Survival", 
                   main = "Yearlings")

str(my.fit)
my.fit

my.fit2 <- survfit(yrlg.ob ~ yrlg.df$Sex)
my.fit2
str(my.fit2)

plot.sex <- plot(my.fit2, xlab="years", log="y",  ylim = c(0.4, 1),xlim=c(0,1),
                 ylab = "survival", main = "Yearlings by sex")
## Cox PH model with sex as a predictor 
cx.yrl <- coxph(yrlg.ob ~ Sex, data = yrlg.df)
cx.yrl
summary(cx.yrl)

#Check PH assumption
cox.zph(cx.yrl)
plot(cox.zph(cx.yrl))

```


From this first model, I think the interpretation is that males and females have different hazards and the hazard ratio is not proportional at the beginning and end of life. It's difficult for me to tell with the plot of cox.zph. The statistics that come out of cox.zph suggest that the hazards are not proportional. It makes sense that this would be true at the beginning and end of life. We can already see that from the KM estimates. Also, the previous analyses I ran on the breeder data set suggested that the sexes do not have different hazards (or not significant...?) and that the hazards are proportional through time. I think this actually makes sense that male and female breeders have about the same risk of dying except for when birds approach the maximum lifespan. Females may senesce earlier but probably don't have enough data to get at this.

So, if I am interpreting the residual plot for diagnosing PH assumption, the hazards are not propotional but it is driven by the beginning and end of life. I think this justifies modeling by stage. Boom. Hope this enough!! 

```{r, echo=FALSE}
bird.df$FldgDate <- as.Date(bird.df$FldgDate, format = "%m/%d/%Y")
bird.df$LastObsDate <- as.Date(bird.df$LastObsDate, format = "%m/%d/%Y")

## Add column to indicate whether or not bird was alive after one year 
## 0 = alive, 1 = dead (same convention as censorship indicator)
bird.df["Censor"] <- 1
## If the date is the April 2016 census date, consider still alive as of then
bird.df$Censor[which(bird.df$LastObsDate == "2016-4-12")] <- 0

## Change back to numeric for survival object 
bird.df$FldgDate <- as.numeric(bird.df$FldgDate)
bird.df$LastObsDate <- as.numeric(bird.df$LastObsDate)

bird.df <- subset(bird.df, bird.df$Days > 0 & bird.df$Yrs > 0)

## Create survival object - only birds that are right censored are still alive
jay.ob <- Surv(bird.df$Yrs, bird.df$Censor, type =c('right'))

## Kaplan-Meier estimate
survfit(jay.ob~1)

my.fit<- survfit(jay.ob~1)

## Plot KM
p1 <- plot(my.fit, main="Kaplan-Meier estimate with 95% CI",log = "y",
           xlab="Years", ylab="surv function (log)",ylim = c(0.001,2))

## KM for sex 
my.fit2 <- survfit(jay.ob ~ bird.df$Sex)
str(my.fit2)

## Plot KM estimate by sex
p2 <- plot(my.fit2, main = "Kaplan-Meier estimate with 95% CI",log = "y",
           xlab="Years", ylab="surv function (log)",ylim = c(0.001,2))

# Cox model assuming time constant covariates 
mod1 <- coxph(jay.ob ~ Sex, data = bird.df)
mod1
summary(mod1)
cox.zph(mod1)
plot(cox.zph(mod1))
```



```{r, echo=FALSE}
prebrdr <- subset(bird.df, Days >= 365)
str(prebrdr)
## 1009 individuals 

## How to the April Census Dates that are the year following fledge year?? 
## Can R even do this? Well, regardless, first step is to convert to dates

## convert dates to date format
prebrdr$FldgDate <- as.Date(prebrdr$FldgDate, format = "%m/%d/%Y")
prebrdr$LastObsDate <- as.Date(prebrdr$LastObsDate, format = "%m/%d/%Y")

## Add column to indicate whether or not bird was alive after one year 
## 0 = alive, 1 = dead (same convention as censorship indicator)
prebrdr["Censor"] <- 1
## If years lived is less than 1, dead
prebrdr$Censor[which(prebrdr$LastObsDate =="2016-4-12")] <- 0

## Change back to numeric for survival object 
prebrdr$FldgDate <- as.numeric(prebrdr$FldgDate)
prebrdr$LastObsDate <- as.numeric(prebrdr$LastObsDate)

pre.ob <- Surv(prebrdr$Yrs, prebrdr$Censor, type =c('right'))

#KM curve 
pre.fit <- survfit(pre.ob~1, conf.type = "log-log")
jay.fit <- plot(pre.fit,log="y", xlim = c(1,14), ylim=c(0.1,1), xlab = "Time (years)", 
                ylab = "Cumulative Survival")
#Km curve by sex
sex.fit <- survfit(pre.ob~ Sex, data = prebrdr)
sex.plot <- plot(sex.fit, log="y", xlim = c(1,14), ylim=c(0.1,1), xlab = "Time (years)", 
                 ylab = "Cumulative Survival")

model1 <- coxph(pre.ob ~ Sex, data = prebrdr)
summary(model1)

cox.zph(model1)
plot(cox.zph(model1))
```

