---
title: "FSJ Surv June Week 4"
author: "Erin Feichtinger"
date: "June 21, 2016"
output: html_document
---
```{r, echo = FALSE}

library(survival)
library(car)
library(coxme)
library(kinship2)
library(SurvRegCensCov)

## Read in April Census Dates
aprilD <- read.csv("April_Census_Dates.csv")
## Convert to date object
aprilD$CensusDate <- as.Date(aprilD$CensusDate, format = "%m/%d/%Y")
## First data table to deal with, known age birds to figure out who died in the first year
## Read in CSV file 
bird.df <- read.csv("Erin_June_FY.csv")
str(bird.df)
## 2371 individuals 
## Change column name to Days for days lived and add a year column
colnames(bird.df)[8] <- "Days"
bird.df["Yrs"] <- bird.df$Days/365.25
## convert dates to date format
bird.df$FldgDate <- as.Date(bird.df$FldgDate, format = "%m/%d/%Y")
bird.df$LastObsDate <- as.Date(bird.df$LastObsDate, format = "%m/%d/%Y")

bird.df["Censor"] <- 1
#Birds that are right censored (indicated by 0) are still alive after 1 year
#post fledgling 
bird.df$Censor[which(bird.df$Yrs > 1)]<-0
#subset to get rid of years less than 0
yrlg.df <- subset(bird.df, bird.df$Yrs > 0)

#change back to numeric for survival object 
yrlg.df$FldgDate <- as.numeric(yrlg.df$FldgDate)
yrlg.df$LastObsDate <- as.numeric(yrlg.df$LastObsDate)

#survival object 
yrlg.ob <- Surv(yrlg.df$Yrs, yrlg.df$Censor, type = c('right'))
my.fit <- survfit(yrlg.ob~1, conf.type = "log-log")
my.fit
str(my.fit)
plot.fit <- plot(my.fit, xlab = "Time (years)",
      log = "y", ylim = c(0.4, 1),xlim=c(0,1), ylab = "Cumulative Survival", 
                   main = "Yearlings")

#By sex
my.fit2 <- survfit(yrlg.ob ~ yrlg.df$Sex)
my.fit2
str(my.fit2)

plot.sex <- plot(my.fit2, xlab="years", log="y",  ylim = c(0.4, 1),xlim=c(0,1),
                 ylab = "survival", main = "Yearlings by sex")
## Cox PH model with sex as a predictor 
cx.yrl <- coxph(yrlg.ob ~ Sex, data = yrlg.df)
cx.yrl
summary(cx.yrl)

#Check PH assumption
cox.zph(cx.yrl)
plot(cox.zph(cx.yrl), main = "First Year")

```

The KM curve for all yearlings shows a slightly curved line with a steeper slope at the very beginning of life which makes sense then begins to level off. During the first year, about half the birds die. 

The KM curves by sex overlap throughout most of the first year except for what looks like a small departure at the beginning and the lines diverge just before one year. 

The output of the Cox model with sex as the only predictor suggests that being male is less risky but the coefficient estimate is not significant. 

The output of the test based on the scaled Schoenfeld residuals suggests that the assumption of proportional hazards is not met. The test for the SexM covariate gives some evidence for non-proportional hazards but it's not strong. However, looking at the plot it seems that the assumption is violated. The function cox.zph uses a smoothing spline for easier interpretation. The dashed lines are +/- 2 standard deviation error envelopes. According to this grey literature example, "systematic departures from a horizontal line are indicative of non-proportional hazards". It's a pretty curvy line for the Cox model with sex as a predictor. So, I guess the hazards are not proportional for the two groups during this time period? 


I did the exact same thing as above for all birds in the data set with no regards to stage. 

```{r, echo=FALSE}
## convert dates to date format
bird.df$FldgDate <- as.Date(bird.df$FldgDate, format = "%m/%d/%Y")
bird.df$LastObsDate <- as.Date(bird.df$LastObsDate, format = "%m/%d/%Y")

## Add column to indicate whether or not bird was alive after one year 
## 0 = alive, 1 = dead (same convention as censorship indicator)
bird.df["Censor"] <- 1
## If the date is the April 2016 census date, consider still alive as of then
bird.df$Censor[which(bird.df$LastObsDate == "2016-4-12")] <- 0

## Change back to numeric for survival object 
bird.df$FldgDate <- as.numeric(bird.df$FldgDate)
bird.df$LastObsDate <- as.numeric(bird.df$LastObsDate)

bird.df <- subset(bird.df, bird.df$Days > 0 & bird.df$Yrs > 0)

## Create survival object - only birds that are right censored are still alive
jay.ob <- Surv(bird.df$Yrs, bird.df$Censor, type =c('right'))

## Kaplan-Meier estimate
survfit(jay.ob~1)

my.fit<- survfit(jay.ob~1)
#summary(my.fit)$surv
## Get information on the KM estimate - returns a list
str(my.fit)

## Plot KM
p1 <- plot(my.fit, main="Kaplan-Meier estimate with 95% CI",log = "y",
           xlab="Years", ylab="survival (log)",ylim = c(0.001,2))

## KM for sex 
my.fit2 <- survfit(jay.ob ~ bird.df$Sex)
my.fit2
str(my.fit2)

## Plot KM estimate by sex
p2 <- plot(my.fit2, main = "Kaplan-Meier estimate with 95% CI",log = "y",
           xlab="Years", ylab="surv function (log)",ylim = c(0.001,2))

# Cox model assuming time constant covariates 
mod1 <- coxph(jay.ob ~ Sex, data = bird.df)
mod1
summary(mod1)

# estimate the distribution of survival times 
p3 <- plot(survfit(mod1), xlab = "years")

#Tests for the PH assumption - tests based on scaled Schoenfeld residuals
cox.zph(mod1)
plot(cox.zph(mod1))

```


Without regard to stage, male birds have a lower risk than females. The KM curve by sex shows that the curves are indistinguishable at the very begining (as we just saw with the F-Y birds). It also appears that the hazards are not proportional over time between the two groups (males and females in this case). I already know from previous analyses sex alone does not predict much for breeder hazard and the hazards are proportional for this group. It's true that I built a few different models with sex, age at first breeding and cohort year as predictors. The model with all three predictors had the lowest AIC score among the models tested. So, knowing all this so far, my prediction is that the helping period is where the most differences in risk and survival probablity are between the sexes. And I suspect that the covariates in the models of each stage are different or have different effects. 

Now on to the next subset of data. I'm having an issue getting the helpers only (I'll explain later) but for this I used birds that became helpers and breeders. I used the same procedure as before by generating survival curves and running a cox model with sex as a predictor. 



The important point is this: there is justification for modeling by stage, with the exception of the breeders, the assumption of proportional hazards is violated. The differences between males and females in hazard is during the period between 1-4 years old. This makes sense biologically. My trouble with the helpers is getting the individuals for the model. I have a plan on how to resolve this however. 


