---
title: "Breeder Survival"
author: "Erin Feichtinger"
date: "September 7, 2016"
output: html_document
---

# Breeder Survival 1981 - 2016

### The data 

These data consist of Florida Scrub-Jay breeders at Archbold Biological Station from 1981 to 2016, i.e. birds who first bred between 1981 and 2016. These data include both birds of known and unknown age. Birds hatched on the study tract are of known age while immigrants are of unknown age. However, there is an estimate of minimum age for unknown age birds. Each row in the data frame represents a record for one individual. There is only one row per individual. The variables include bird ID, sex, ageknown (1 or 0), minimum age at first breeding, year of first breeding, years spent breeding, last date observed and censorship indicator. The right censored birds are the ones that are still alive as of the April 2016 monthly census. 

N = 958 birds
N males = 480
N females = 479

### Model types
1. Kaplan-Meier nonparametric survival estimates
2. Time-constant Cox Proportional Hazards Models
3. Accelerated Failure Time Models 
4. Frailty models
    + Cox mixed effects models
    + AFT frailty with gamma distribution 


#### Kaplan Meier Survival Curves 
```{r, echo=FALSE}
library(survival)
library(coxme)
library(plyr)
brd <- read.csv("Breeders.csv")
brd <- brd[!duplicated(brd),]
colnames(brd)[1] <- "ID"
colnames(brd)[2] <- "Band"
colnames(brd)[3] <- "FY"
colnames(brd)[4] <- "Fbreed"
brd <- brd[,c(1,2,8,7,6,3,4,5)]
colnames(brd)[5] <- "MinAgeFBr"
brd[brd==""] <- NA
brd <- na.omit(brd)
brd$Fbreed <- as.Date(brd$Fbreed, format ="%m/%d/%Y")
brd$LastObsDate <- as.Date(brd$LastObsDate, format = "%m/%d/%Y")
brd["Yrs"] <- (brd$LastObsDate - brd$Fbreed)/365.25
brd$Yrs <- round(brd$Yrs, digits = 1)
brd["Censor"] <- 1
brd$Censor[which(brd$LastObsDate =="2016-4-12")] <- 0
brd <- subset(brd, brd$Yrs > 0)
brd <- subset(brd, brd$FY >= "1981")
brd$Fbreed <- as.numeric(brd$Fbreed)
brd$LastObsDate <- as.numeric(brd$LastObsDate)
brd$Yrs <- as.numeric(brd$Yrs)
brd.ob <- Surv(brd$Yrs, brd$Censor,type= c('right'))
brd.fit <- survfit(brd.ob ~ 1, conf.type = "log-log")
kmplot <- plot(brd.fit, xlab="Breeding time span (years)", log = "y", 
               ylab = "Cumulative Survival (log)", main = "Breeders",
               ylim = c(0.01,1), xlim = c(0,15))

brd.sex <- survfit(brd.ob ~ brd$Sex, conf.type="log-log")
sxplot <- plot(brd.sex, col = c("darkblue","darkorange3"), 
               xlab = "Breeding time span (years)", log="y", 
               ylab = "Cumulative Survival (log)", main = "Breeders",
               ylim = c(0.01,1), xlim = c(0,15))
legend("topright", c("Females","Males"), col=c("darkblue","darkorange3"),
       lty = c(1,1),lwd=1)
```


The survival curves show the cumulative survival over the breeding lifespan. The x axis starts at zero, meaning the start of the breeding interval. Birds start breeding at different ages so this time scale is not age. However, we can assume that most birds are at least 2 by the time they breed for the first time, so time point zero here corresponds to approximately 2. For example, at time point 2, that could represent birds of ages 4+.   
Half of all breeders that started the interval are gone by a breeding span of 3 years. 80% of birds that started the interval have disappeared by 8 years of breeding (correspoding to about 10 years of age at a minimum). Most breeders have disappeared by a breeding timespan of 10 years (about 12 years of age).  
The estimated survival probability for males and females is similar until about 6 years into the breeding period (at least 8 years old), then the curves start diverging. There is evidence of senecense in this species (MacDonald 1996; Fox et al. 2006) and there may be something related to this and females that lowers survival probability. However, very few birds make it past 10 years of age (time point 12 represents 14 years of age, approx) so we do not have the data to distingiush survival differences at old ages.   

Summary stats breeding lifespan 
Birds  | Mean (yrs) | SD
-------|------------|-----
All    | 3.39       | 2.96
Females| 3.37       | 2.84
Males  | 3.41       | 3.07

Summary stats (minimum) age at first breeding
Birds  | Mean (yrs) | SD
-------|------------|-----
All    | 2.58       | 1.02
Females| 2.53       | 0.95           
Males  | 2.63       | 1.08    

The average breeding lifespan and age at first breeding are similar for males and females. There is a lot of variation in how long birds spend breeding. The standard deviation is almost as big as the mean. The mean minimum age at first breeding is similar between males and females, along with the standard deviation. This suggests that most birds are at least 2 or 3, on average, when breeding for the first time. 

#### Cox Proportional Hazard Models 
I estimated the null model LL by using an intercept only in the first Cox model (LL = -4863.76). Then, I added sex as a predictor. I assume that the hazards are proportional, since all individuals are likely to experience increases or decreases in hazard at the same time (e.g. increase in predators, other stochastic events). The R output for this model is below. 

```{r, echo = FALSE}
cx1 <- coxph(brd.ob ~ Sex, data = brd)
summary(cx1)
```


The coefficient estimate for sex(male) is negative, meaning males have about a 6% reduction in hazard compared to females, but it is not significant. The loglikelihood ratio tests indicated that this is not a very good model given the data. Sex does not add much predictive power to the model.  
We can test the proportional hazards assumption of this model. 
```{r, echo=FALSE}
plot(cox.zph(cx1))
```

A straight line at zero indicates proportional hazards through time. This is mostly met except for the end. From this, I think the hazards are approximately proportional. 

The next Cox PH model has (minimum) age at first breeding as a predictor. 

```{r, echo=FALSE}
cx2 <- coxph(brd.ob ~ MinAgeFBr, data = brd)
summary(cx2)
```

This model is a slightly better fit than the null model or the model with sex. The coefficient for minimum age bred is positive, meaning there is an increased hazard (about 9%) each year delayed breeding (I think, that's what Bruce said). The assumption of proportional hazards is met here.  
However, we don't know if this is causal. As in, does breeding for the first time at an older age make your life riskier, or it is just because breeding at an older age means breeding closer to the end of the lifespan (closer to death). 

The 3rd Cox model is with year (coded as factor) as a predictor. This is the year each individual became breeders for the first time. 

```{r, echo=FALSE}
cx3 <- coxph(brd.ob ~ FY, data = brd)
cx4 <- coxph(brd.ob ~ Sex + MinAgeFBr, data=brd)
cx5 <- coxph(brd.ob ~ Sex + MinAgeFBr + FY, data=brd)
cx6 <- coxph(brd.ob ~ Sex + MinAgeFBr + FY + Sex*MinAgeFBr, data = brd)
```

The warning message says that beta may be infinite. I'm guessing this is a problem of convergence and the beta estimate is biased...? I will have to ask about this. None of the coefficient estimates are significant. This is puzzling to me. In fact, the AIC scores say this is the least best fitting model of all three, however, the LL ratio test is the highest for this one. I think the frailty models will be more informitive. Either way, none of these simple models are very good given the data. Three more models fit, all three predictors no interaction, all 3 plus interaction between age and sex and age and sex.

Model                     | AIC     | df
--------------------------|---------------
Sex                       | 9728.7  | 1
Age                       | 9722.9  | 1
Year                      | 9744.6  | 35
Sex + Age                 | 9723.7  | 2
Sex + Age + Year          | 9744.3  | 37
Sex + Age + Year + Sex*Age| 9740.5  | 38

According to AIC, the best model of these 6 is the model with minimum age at first breeding. The next best model is sex and age. Year doesn't seem to explain much variation. All models with year are worse fits than those without, at least for this set and according to AIC. 

#### Accelerated Failure Time Models 

#### Frailty Models

##### Mixed effects Cox models 

The first model has minimum age of first breeding as the fixed effect and year of first breeding as the random effect. Here, we are interested in the variance of the random effect. 

```{r, echo=FALSE}
frail1 <- coxme(brd.ob ~ MinAgeFBr + (1|FY), data = brd)
summary(frail1)
```

Interestingly, the variance of the random effect is not very high. This is opposite to what Gordon found in the 2006 paper. Did I do this correctly? This is really bizzare. The analysis of deviance tells me that the year Cox PH model has the largest reduction in the LL. Not sure what to make of this. Is it because I have 2016 included? What happens if I take that out? 

```{r, echo=FALSE}
dev.compare <- anova(cx1, cx2, cx3, test="Chisq")
dev.compare
```

I don't know how Gordon estimated a model deviance, but I am unable to find anything that gives one number out in R like he reported in the paper. I can compute the analysis of deviance table and it shows me the LL when each term is added. When I compare the models with sex, age, and year, respectively, the lowest LL is for model 3. However, the AIC value is the highest out of the set of models and none of the coefficients for the 35 years are significant. I really don't know what to make of this. 

I will have to talk to Gordon (and probably do a lot of searching for answers on my own) about this whole deviance business and how they got the reported values in their paper. I don't see any "deviance" term. Just the loglikelihoods and the reduction in those. Is this the same thing? 





