---
title: "FSJ Adults"
author: "Erin Feichtinger"
date: "November 4, 2016"
output: pdf_document
---
#Survival of adult Florida Scrub-Jays in a stage based population 

Florida Scrub-Jays can be in one of several stages throughout their lives. Beyond the nestling and fledgling stages, birds are classified as juveniles (day 75 to 1 yr old), prebreeders/helpers (adult), and breeders. The last two stages are also age-structured in that birds do not disperse or become breeders at the same age. The average age of first breeding is two, but there are several first time breeders who are 3 years old. There are instances of birds becoming first time breederes at older ages, but this is rare. On average, by age 3, birds have either disappeared or became a breeder.  

Due to the nature of the population structure, patterns of survival could be different across the stages, especially for hatch year birds and prebreeders due to biology and dispersal behavior. Furthermore, the underlying causes of variation in survival could be different between stages. For example, young birds are vulernable to predation and rely on their parents and any helpers to detect predators. Prebreeders stay in (usually) their natal territories until dispersing to establish a breeding territory. Females tend to disperse farther on average and thus could be exposed to greater risk than prebreeding males seeking a territory. 

Due to this dispersal pattern, especially of females, the methods typically used to analyze survival are inadequate to esimate the risk of mortality for prebreeders because we cannot distinguish between death or dispersal. There is nearby scrub that is outside the monitored area in which birds can disperse. Most birds in the ABS population have a last observed date and while this usually indicates death, it could mean that bird was last observed on the study tract but is alive elsewhere.  

Using Cox Proportional Hazards models to analyze risk of mortality in the pre-breeder period proved inadequate. The estimated coefficient of hazard for males suggested a 40 % decrease in hazard relative to females during this period. Furthermore, the proportion of birds that started the stage as a prebreeder and transitioned to breeder, was estimated to be 70% for males and 30% for females. The breeder sex ratio is approximately 1:1 so there would have to be a large number of females dispersing into the study tract from elsewhere. Of the unknown age birds (N = 290), 60% are females (N=174).  

Bayesian methods may be able to shed light on this problem using a priori information on dispersal patterns and population genetics in Florida Scrub-Jays. A recent paper described such methods for distinguishing between dispersal and death for African lions, a system in which males disperse from their maternal prides. 

For these reasons, we are analyzing survival in two of the life stages of Florida Scrub-Jays, hatch year birds (fledge to 1 yr) and breeders. A bird is classified as a breeder when it is known to lay or sire an egg for the first time. This stage is also age-structured. 


##Florida Scrub-Jay Breeder Survival 

These data include all breeders, of known and unknown age, residing in monitored territories at Archbold Biological Station from 1981 to 2016. What is the pattern of survival during the breeding stage among years? How does survival probability of breeders vary between individuals both within and among years? What are the effects of territory, social group and family identity on survival? Is breeder survival different between males and females? Are birds who start breeding at an older age inherently leading riskier lives because of the status of an older than average breeder, or are they simply closer to the end of the expected lifespan? 

```{r, echo=FALSE}
library(survival)
library(ggplot2)
library(car)
library(coxme)
library(kinship2)
library(SurvRegCensCov)
library(parfm)
library(plyr)

brd <- read.csv("Breeders.csv")
brd <- brd[!duplicated(brd),]

colnames(brd)[1] <- "ID"
colnames(brd)[2] <- "Band"
colnames(brd)[3] <- "FY"
colnames(brd)[4] <- "Fbreed"

brd <- brd[,c(1,2,8,7,6,3,4,5)]
colnames(brd)[5] <- "MinAgeFBr"

brd[brd==""] <- NA
brd <- na.omit(brd)

brd$Fbreed <- as.Date(brd$Fbreed, format ="%m/%d/%Y")
brd$LastObsDate <- as.Date(brd$LastObsDate, format = "%m/%d/%Y")

brd["Yrs"] <- (brd$LastObsDate - brd$Fbreed)/365.25
brd$Yrs <- round(brd$Yrs, digits = 1)
brd["Censor"] <- 1
brd$Censor[which(brd$LastObsDate =="2016-4-12")] <- 0

brd <- subset(brd, brd$Yrs > 0)
brd <- subset(brd, brd$FY >= "1981")
brd$FY <- as.factor(brd$FY)
brd$Fbreed <- as.numeric(brd$Fbreed)
brd$LastObsDate <- as.numeric(brd$LastObsDate)
brd$Yrs <- as.numeric(brd$Yrs)
brd.ob <- Surv(brd$Yrs, brd$Censor,type= c('right'))
brd.fit <- survfit(brd.ob ~ 1, conf.type = "log-log")
year.fit <- survfit(brd.ob ~ brd$FY, conf.type="log-log")
brd.sex <- survfit(brd.ob ~ brd$Sex, conf.type="log-log")


males <- subset(brd, Sex == "M")
females <- subset(brd, Sex == "F")

cat("No. of Known Age Birds = ", nrow(brd[brd$AgeKnown == "Y", ]))
cat("No. of Unknown Age Birds = ", nrow(brd[brd$AgeKnown == "N", ]))

cat("No. of Females = ", nrow(brd[brd$Sex == "F", ]))
cat("No. of Males = ", nrow(brd[brd$Sex == "M", ]))

cat("No. of unknown age females = ", nrow(females[females$AgeKnown == "N", ]))
cat("No. of unknown age males = ", nrow(males[males$AgeKnown == "N", ]))

cat("No. of known age females = ", nrow(females[females$AgeKnown == "Y", ]))
cat("No. of known age males = ", nrow(males[males$AgeKnown == "Y", ]))

ggplot(brd, aes(MinAgeFBr, fill = Sex)) +
  geom_bar(position="dodge") +
  xlab("Minimum Age at First Breeding") +
  ylab("Frequency") +
  labs(title = "Distribution of Ages at First Breeding")

```
## Main findings 

The total sample size N = 958 (478 females and 470 males) with 668 known age individuals and 290 of unknown age. Of those 290, 174 are females and 116 are males. Of the known age birds, 304 are females and 364 are males. This is expected because female Scrub-Jays disperse farther, on average, than males. We would expect there to be more unknown age females than males and more known age males than females. The difference between males and females in both groups is 58 and 60, so it seems that there are as many coming in than leaving.  

I initially estimated the Kaplan-Meier estimates (non-parametric) for survival for all birds together, by sex and by year to explore the pattern of cumulative survival over time. The KM curves for males and females overlapp until the end of the maximum lifespan, although the sample size for this period is very small since few birds make it past age 10. 

```{r, echo=FALSE}
kmplot <- plot(brd.fit, xlab="Breeding time span (years)", log = "y", 
               ylab = "Cumulative Survival (log)", main = "Breeders",
               ylim = c(0.01,1), xlim = c(0,15))

yrplot <- plot(year.fit, xlab ="Breeding time span (years)", log = "y",
               ylab = "Cumulative Survival (log)", main = "Breeders", 
                ylim = c(0.01,1), xlim = c(0,15))

sxplot <- plot(brd.sex, conf.int=TRUE, col = c("darkblue","darkorange3"), 
               xlab = "Breeding time span (years)", log="y", 
               ylab = "Cumulative Survival (log)", main = "Breeders",
               ylim = c(0.01,1), xlim = c(0,15))
legend("topright", c("Females","Males"), col=c("darkblue","darkorange3"),
       lty = c(1,1),lwd=1)

```

In these plots of the Kaplan-Meier survival estimates, the time period starts when birds become breeders, thus the x-axis is the time spent as a breeder starting at t = 0, the time point in which an individual becomes a breeder. Birds start breeding at different ages, so age is not represented by the x axis. However, most birds start breeding at age 2 or 3, so you can add that to approximate the ages of the birds. This isn't immediately obvious because cumulative survival is in a log scale, but about 50% of the breeders have disappeared around 3 years of breeding, or approximately age 5 or 6. 

As far as our methods can detect with the data that we have, male and female survival is not different in the breeder stage. There may be differences in survival near the end of the expected and maximum lifespan, but we have too few observations at those ages. 

```{r}
cx1 <- coxph(brd.ob ~ Sex, data = brd)
summary(cx1)
anova(cx1)

cx2 <- coxph(brd.ob ~ MinAgeFBr, data = brd)
summary(cx2)
anova(cx2)

cx3 <- coxph(brd.ob ~ FY, data = brd)
anova(cx3)
cx4 <- coxph(brd.ob ~ Sex + MinAgeFBr, data=brd)
anova(cx4)

cx5 <- coxph(brd.ob ~ Sex + MinAgeFBr + FY, data=brd)
anova(cx5)
cx6 <- coxph(brd.ob ~ Sex + MinAgeFBr + FY + Sex*MinAgeFBr, data = brd)
cx6
anova(cx6)

dev.compare <- anova(cx1, cx2, cx3, cx4, cx5, cx6, test="Chisq")
dev.compare
```

I don't know what the interaction term even means. Not sure if this is correct. 

I am working on buidling time-varying Cox models. 
