---
title: "Survival Update"
author: "Erin Feichtinger"
date: "February 23, 2016"
output: pdf_document
---

This document is a continuation and update of the report titled "Florida Scrub Jay Survival" knitted on February 21, 2016. Some of the information is the same but there are some updates and I cut out the part on the "practice" data set where the breeders had only one row per individual. The data set used here is for breeders and each bird can have multiple rows in the data frame for each year bred. 
```{r, echo = FALSE}
#use this on school desktop
setwd("C:/Users/efeichtinger/LittleBlueDinos")
library(survival)
library(car)
library(kinship2)
library(SurvRegCensCov)

##Read in CSV file of male and female breeders with mulitple rows for each bird
bird.df <- read.csv("Erin_Breeders_All_Years.csv")

#remove duplicates  - for years where there was more than one nest in a year
jay.df<- bird.df[!duplicated(bird.df),]

colnames(jay.df)[1] <- "ID"
colnames(jay.df)[2] <- "Band"

#convert dates to date format
jay.df$MinDate <- as.Date(jay.df$MinDate, format = "%m/%d/%Y")
jay.df$LastObsDate <- as.Date(jay.df$LastObsDate, format = "%m/%d/%Y")

#subtract dates to get number of days
date.diff<- jay.df$LastObsDate-jay.df$MinDate

#and survival period in years, account for leap year 
jay.df["Yrs"] <- date.diff/365.25

#very important piece of code for the model to work properly, remove any 
#weird entries like birds that have negative years of experience or a negative
#survival interval 
jay.df <- subset(jay.df, jay.df$YrsExp >= 0 & jay.df$Yrs > 0)
str(jay.df)

#add column for censorship status, in survival package - 0=alive, 1=dead
jay.df["censorship"] <- 1
#If last observed date = 10/14/2015, 0 for still alive today
jay.df$censorship[which(jay.df$LastObsDate=="2015-10-14")]<-0

#change back to numeric for survival object 
jay.df$MinDate <- as.numeric(jay.df$MinDate)
jay.df$LastObsDate <- as.numeric(jay.df$LastObsDate)
jay.df$Yrs <- as.numeric(jay.df$Yrs)

```

str() tells us what is in the data frame and we can see that each bird has an id, the minimum date known to have laid or sired an egg (MinDate), date last observed, age at first breeding, sex, and years of breeding experience among a few other things. I still have to add in group size. There is a query already built in the data base that gives the count of birds in a territory but the challenge is figuring out how to connect it to my query where the data at hand came from. I'm working on it.  

```{r}
#Create survival object - IS THIS CORRECT?? 
jay.ob <- Surv(jay.df$Yrs, jay.df$censorship, type =c('right'))
jay.lifetab <- survfit(jay.ob~1, conf.type = "log-log")
jay.fit <- plot(jay.lifetab, xlab = "Time (years)", 
          ylab = "Cumulative Survival", main = "FL Scrub Breeder survival")

#Grouping by sex - following example "Cox Regression in R" J. Fox
km.sex <- survfit(jay.ob ~ jay.df$Sex, conf.type = "log-log")
km.fit <- plot(km.sex, xlab = "Time (years)", 
               ylab = "Survival", main = "Survival by Sex")
```

Two K-M curves where the first one is all breeders and the second figure splits the birds up by sex. I couldn't figure out to label which line is which as of right now (11:45 AM...mmmmm lunch). 

```{r}
males <- subset(jay.df, jay.df$Sex == "M")
females <- subset(jay.df, jay.df$Sex == "F")
#table of means & SD for males and females 
means <- cbind(mean(females$Yrs), mean(males$Yrs))
sd <- cbind(sd(females$Yrs), sd(males$Yrs))
colnames(means) <- c("Females", "Males")
m.sd <- rbind(means, sd)
rownames(m.sd) <- c("Mean", "SD")
m.sd
```

A table with the mean and standard deviation of breeding lifespan in years for males and females.  

Now, what is the relationship between age and years of breeding experience? This simple plot could help. It's years of breeding experience as a function of age. 

```{r}
#simple plots to look at age and years of breeding experience
plot(jay.df$CurrentAge, jay.df$YrsExp, xlab = "Current Age",
     ylab = "Years Experience")
```

Well, age at first breeding goes from 1 to 10. I think most birds breed for the first time much younger than 10, but nonetheless there is variation in the age at first breeding. Can we look at these two things separetly? My first thought is "yes" but I'm only about 60% confident on that.  

Now let's try fitting some simple Cox PH models with 1, 2, and 3 covariates.

```{r}
#First Cox Models
#Years of breeder experience as predictor
cox1 <- coxph(jay.ob ~ YrsExp, data = jay.df)
summary(cox1)
#Check for violation of proportional hazard 
res.cox1 <- cox.zph(cox1)
res.cox1

#Sex as predictor
cox2 <- coxph(jay.ob ~ Sex, data = jay.df)
summary(cox2)
#Age as predictor
cox3 <- coxph(jay.ob ~ CurrentAge, data = jay.df)
summary(cox3)

#Include years experience, age, and sex, no interactions
cox4 <- coxph(jay.ob ~ YrsExp + Sex + CurrentAge, data = jay.df)
summary(cox4)

#Check that hazards are proportional 
res.cox4 <- cox.zph(cox4, transform = "km")
res.cox4
plot(res.cox4)

extractAIC(cox1)
extractAIC(cox2)
extractAIC(cox3)
extractAIC(cox4)
```

I'm not sure how to interpet the output and the figure here. I followed an example that I found online but it didn't have anything in the way of interpretation. I can't tell from looking at this if the hazards are proportional or not.  

For the Cox models, I can't tell if these are okay or really bad models given the data. From extractAIC(), the cox model with 3 covariates has the lowest AIC value of these four models, but I don't know how to tell if these are even acceptable models. 

```{r}
#AFT model with exponential distribution and years of experience 
AFT.exp1 <- survreg(jay.ob ~ YrsExp, data = jay.df, dist = "exponential")
#AFT model with Weibull distribution and years of experience
AFT.weibull <- survreg(jay.ob ~ YrsExp, data = jay.df, dist = "weibull")
summary(AFT.exp1)
summary(AFT.weibull)

#AFT model with exponential distribution and sex
AFT.exp2 <- survreg(jay.ob ~ Sex, data = jay.df, dist = "exponential")
#AFT model with Weibull distribution and sex
AFT.weibull2 <- survreg(jay.ob ~ Sex, data = jay.df, dist = "weibull")
summary(AFT.exp2)
summary(AFT.weibull2)

#AFT model with exponential distribution and sex
AFT.exp3 <- survreg(jay.ob ~ CurrentAge, data = jay.df, dist = "exponential")
#AFT model with Weibull distribution and sex
AFT.weibull3 <- survreg(jay.ob ~ CurrentAge, data = jay.df, dist = "weibull")
summary(AFT.exp3)
summary(AFT.weibull3)

#All 3 covariates
AFT.exp4 <- survreg(jay.ob ~ YrsExp + Sex + CurrentAge, data = jay.df,
                    dist = "exponential")
AFT.weibull4 <- survreg(jay.ob ~ YrsExp + Sex + CurrentAge, data = jay.df,
                        dist = "weibull")

#Compare AFT models with Cox PH model that has 3 covariates 
summary(cox4)
summary(AFT.weibull4)
summary(AFT.exp4)

```

Now let's talk carefully about this output. 
